---
title: "Basics GO analysis with lichengo"
output: 
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 4
vignette: >
  %\VignetteIndexEntry{basics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

## lichengo

`lichengo` is a package for generating gene ontology (GO) analysis for using genes of lichen-forming species.

## Load package and data

Load package.

```{r setup}
library(lichengo)
```

Check available assemblies.

```{r check-assemblies}
available_assemblies()
```
Select assembly for the species of interest.

```{r select-assembly}
assembly = 'treb_jgi'
```

Retrieve the list of genes for teh selected species. 

```{r all-genes}
genes <- all_genes(assembly)
genes %>% head()
```

## Load expression data

Import your expression data including - at least - the gene IDs, the log fold changes and the $p$-values. 

The `lichengo` package offers an example of expression data for the available lichen-forming organisms. 

```{r expression-file, message=FALSE}
filename = paste(assembly, 'expression.csv', sep = '_')
file <- system.file("extdata", filename, package = "lichengo", mustWork = TRUE)
file
```

You can do that with `read_csv()`. Minimally it should contain the gene ID, the log fold change and the $p$-value.

```{r expression-data}
expression <- readr::read_csv(file)
head(expression, 4) 
```

### Filter as required

You can now filter the genes to select only the ones with e.g $p <= 0.05$, using `filter()`

```{r filter-data, message=FALSE}
filtered_expression <- filter(expression, p.adj <= 0.05)
filtered_expression %>% head(., 4)
```

### Extract the gene ID

You can now extract the `gene_id` column using the `$` syntax

```{r, message=FALSE}
selected_gene_ids <- filtered_expression$gene_id
head(selected_gene_ids, 4)
```


## GO enrichement

The GO enrichment is done in the `lichengo` package. Load that and use the `enrich()` function, passing it the vector of `selected_gene_ids` to calculate the enrichment.
adjusted pvalue cutoff `pval` on enrichment tests to report. 

```{r}
enrichment <- enrich(genes = selected_gene_ids, 
  assembly = assembly, 
  pval = 0.05)

result_table <- as.data.frame(enrichment)
head(result_table)
```

## Visualisation

Visualisations can be done with the `enrichplot` package.

```{r}
library(enrichplot)
```

### Barplot

```{r}
barplot(enrichment, x = "GeneRatio", showCategory = 5)
```


```{r}
mutate(enrichment, qscore = -log(p.adjust, base = 10)) %>% 
    barplot(x = "qscore")
```

### Dotplot

```{r}
dotplot(enrichment, showCategory = 5)
```

